---
title: "Introduction to ramify"
author: "Brandon M. Greenwell"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette: 
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Introduction to ramify}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Matrices are a fundamental component of any scientific programming language. In fact, MATLAB, a scientific language used by millions of engineers, is an acronym for MATrix LABoratory. In R, there are several ways to create and manipulate matrices. However, R's matrix syntax is quite different from most other programming languages --- most of which do share a familiar syntax. In this vignette, we discuss `ramify`, a simple package providing R with additional matrix functionality. The added functions and syntax should make R users who are more familiar with other languages --- such as [Julia](http://julialang.org/), [MATLAB](http://www.mathworks.com/products/matlab/)/[GNU Octave](https://www.gnu.org/software/octave/), or [Python](https://www.python.org/) --- feel closer to home.

## Introduction

There are several methods available for constructing matrices in R. The usual approach is to use the base function `matrix`. For example, the following snippet of code creates the matrix 
$\bigl(\begin{smallmatrix} 1 & 2 & 3 & 4 \\ 5 & 6 & 7 & 8 \end{smallmatrix} \bigr)$:

```{r}
  matrix(c(1, 2, 3, 4, 5, 6, 7, 8), nrow = 2, byrow = TRUE)
```

Notice that we need to specify the shape of the matrix (here we used `nrow = 2`, but could also have set `ncol  = 4`, or both). Also, by default, R fills matrices using column-major order. To fill the matrix using row-major order, we had to set the`byrow` argument in `matrix` to `TRUE`.(Even though we set `byrow = TRUE`, the matrix is still stored in column-major order. This may have unintended side-effects. For instance, removing the dimension attribute flattens the matrix into a vector by appending each column together.) Although the code is simple, users coming from other scientific languages may prefer a more familiar syntax. 

The following table gives some examples for constructing matrices in five of the most popular scientific languages.

Language          | Syntax                                | Note
----------------- | ------------------------------------- | ---------------------
Julia             | `[1 2 3 4; 5 6 7 8]`                  | NA   
Mathematica       | `{{1, 2, 3, 4}, {5, 6, 7, 8}}`        | NA   
MATLAB/GNU Octave | `[1, 2, 3, 4; 5, 6, 7, 8]`            | commas may be omitted
Python            | `[[1, 2, 3, 4], [5, 6, 7, 8]]`        | NA   
Python+NumPy      | `numpy.mat("1, 2, 3, 4; 5, 6, 7, 8")` | commas may be omitted

What is most convenient about the matrix syntax expressed in column two is the visual separation of rows. This makes it quite easy for the user to see the structure of the matrix they are working with. You do not see this with the traditional matrix function in R, unless you force individual rows onto new lines manually:

```{r}
matrix(c(1, 2, 3, 4,
         5, 6, 7, 8), nrow = 2, byrow = TRUE)
```

But this is rather hacky in comparison and still requires the user to specify how the vector should be split up (e.g., `nrow = 2` and `byrow = TRUE`). The `ramify` package's main function `mat` extends `matrix` by adding this simplicity using the more common syntax.

The `ramify` package is hosted on GitHub at \url{https://github.com/bgreenwell/ramify}. It is also available on CRAN. To install the latest stable release from CRAN:

```{r, eval=FALSE}
install.packages("ramify")  # latest stable release
```

To install the development version from GitHub you can use the [`devtools`](https://cran.r-project.org/package=devtools) package:

```
# install.packages("devtools")
devtools::install_github("bgreenwell/ramify")  # development version
```

Bug reports or issues should be submitted to \url{https://github.com/bgreenwell/ramify/issues}. Suggestions for improvement can be emailed directly to the package maintainer. The version of `ramify` used in this paper is version 0.3.1.

## The `mat` function

The main function in this package is `mat`. For all intents and purposes, `mat` is simply an extension of the base function `matrix` that adds two new S3 methods: a method for class `"character"` and another for class `"list"`. Fortunately, since `mat` is simply a wrapper around `matrix`, we can still use it in the exact same way. That is, any use of `matrix` also applies to `mat`.

### Character method
Function `mat` provides a new way of creating matrices using a convenient string initializer (not unlike the matrix constructor in NumPy). For instance, we can recreate the matrix $\bigl(\begin{smallmatrix} 1 & 2 & 3 & 4 \\ 5 & 6 & 7 & 8 \end{smallmatrix} \bigr)$ using `mat` as follows:

```{r}
library(ramify)  # load package
mat("1, 2, 3, 4; 5, 6, 7, 8")
```

The colon operator can also be used, as in `mat("1:4; 5:8")`.

The character method of `mat` is very simple. First, the character string is split on the semicolons creating character strings representing the row vectors. Second, the resulting character strings are further split on commas. The individual characters are then parsed, evaluated, and fed to `matrix`. 

The first argument to `mat` (and the only one required) should be a character string in which semicolons separate row vectors and commas separate individual elements. However, at the time of this writing, there are two optional arguments: `rows` and `sep`. `rows` accepts a logical indicating whether the semicolon separates rows (`rows = TRUE`) or columns (`rows = FALSE`). The default is `TRUE`. For instance, to create the matrix 
$$
  \begin{pmatrix} 1 & 5 \\ 2 & 6 \\ 3 & 7 \\ 4 & 8 \end{pmatrix}
$$
just write

```{r}
mat("1:4; 5:8", rows = FALSE)  # ; separates columns
```

The second optional argument, `sep`, accepts a character vector containing regular expressions to use for splitting up the individual elements within each row/column. By default, `sep = ","`. To change the default behavior of separating individual elements by commas, change the value of `sep` to any other valid character. For example, in order to use spaces instead of commas, write 

```{r}
mat("1 2 3 4; 5 6 7 8", sep = "")  # blank spaces separate columns
```

To bypass setting these options every time, the user can change them globally with 

```{r}
options(mat.rows = FALSE, mat.sep = "")  # change default behavior
```

R functions can be used within the character string as well, but one must be careful. For example, `mat("rnorm(10)")` works because there are no semicolons or commas, hence, the character string is just parsed and evaluated. However, `mat("rnorm(10, sd = 3)")` produces an error because `mat` will split the character string on the comma resulting in two substrings that cannot be parsed:

```{r}
strsplit("rnorm(10, sd = 3)", split = ",")
```

One particular way around this is to set the `sep` option to `NULL`:

```{r}
mat("rnorm(5); rnorm(5, mean = 10)", sep = NULL)
```

\subsection{List method}
There is often a need to construct matrices from the elements of a list. For example, I tend to store the results of simulations in a list and later want to treat the elements (usually a vector of results) as the rows/columns of a matrix. While this is not difficult to do manually, I frequently have to stop and think for a minute of the best way to accomplish this. 

For example, suppose we want to take the list

```{r}
z <- list("a" = 1:10, "b" = 11:20, "c" = 21:30)
```

and convert it into a matrix of the form
$$
  \begin{pmatrix} 1  & 2  & 3  & ... & 10 \\ 
                  11 & 12 & 13 & ... & 20 \\ 
                  21 & 22 & 23 & ... & 30\end{pmatrix}
$$
Three approaches come to mind:

```{r}
# Approach 1
matrix(unlist(z), nrow = 3, byrow = TRUE)

# Approach 2
do.call(rbind, z)

# Approach 3
t(simplify2array(z))
```

All three approaches succeed in constructing the correct matrix. Using `matrix` is simple, but requires the user to flatten the list first and specify the number of rows (or, equivalently, the number of columns) ahead of time. Approach two is probably the best, but novice users are not likely to be familiar with `do.call` or the `rbind` and `cbind` method of combing vectors in R. Similarly, in the third approach, new users are not likely to be familiar with `simplify2array`, and the user has to take the transpose of the resulting matrix.

Using `mat` we can construct the matrix as follows:

```{r}
mat(z) 
```

Notice how the element names are preserved as row names. We could force the list elements to be columns instead by setting `rows = FALSE`. Similar to approach two, `mat` uses `do.call` with `rbind` and `cbind` to construct matrices from lists.

### Better printing for "large" matrices

Printing matrices in R can be troublesome. By default, R will try to print the entire matrix to the screen until it reaches its limit (given by `getOption("max.print")` which is usually set to 10000). Columns will also spill onto multiple rows, if there are enough of them. It is therefore not very useful to print even moderately large matrices.

Advanced users typically use the `head` and `tail` functions for viewing, respectively, the first and last few rows of a matrix or data frame. However, if there are a lot of columns, they will still spill over onto the next row on the screen. To see this, try running the following 

```{r}
m <- matrix(rnorm(1000000), nrow = 1000)
# head(m)
```

A new generic function `pprint` (which stands for pretty print) is included with the package. S3 methods for objects of class `"matrix"` and `"data.frame"` are also available. Since its a generic, users can add new methods to suit their needs (e.g., special printing for lists). 

With `pprint`, large matrices are printed in a much nicer way. For instance,

```{r}
m <- randn(1000, 1000)  # see Table 2 for a description of randn
pprint(m)
```

produces a $1000 \times 1000$ matrix of normally distributed random numbers, but only the first few rows and columns, along with the last row and column, of the resulting matrix are shown. The dimension and storage mode are also printed above the matrix.

This printing behavior is also useful for viewing large numeric data frames. `pprint` will convert data frames to a matrix via the base function `data.matrix`. Consequently, logicals and factors will be converted to integers before being printed. For data frames, nicer printing is available via the [`"tbl_df"`](https://cran.r-project.org/package=dplyr) class from the popular `dplyr` package.

## Data frames and block matrices

Package `ramify` also provides two functions similar to `mat`, namely, `dmat` and `bmat`. Function `dmat` works exactly the same as `mat` but results in a data frame, rather than a matrix. The `bmat` function constructs block matrices using a character string initializer.

### Making data frames with dmat
Converting a list to a data frame in R is rather simple. An example is given in the code snippet below. 

```{r}
# List holding individual variables
z1 <- list(Height = c(Joe = 6.2,   Mary = 5.7,   Pete = 6.1),
           Weight = c(Joe = 192.2, Mary = 164.3, Pete = 201.7),
           Gender = c(Joe = 0,     Mary = 1,     Pete = 0))

as.data.frame(z1)  # convert z1 to a data frame
```

When applied to a list, `as.data.frame` treats the individual elements (i.e., `Height`, `Weight`, and `Gender`) as columns. However, it is often the case that list elements represent records, rather than individual variables. (This is a common use for Python dictionaries which are naturally imported into R as a list.) For example, consider the same list, but in a different format:

```{r}
# List holding records (i.e., individual observations)
z2 <- list(Joe  = c(Height = 6.2, Weight = 192.2, Gender = 0),
           Mary = c(Height = 5.7, Weight = 164.3, Gender = 1),
           Pete = c(Height = 6.1, Weight = 201.7, Gender = 0))
```

Here, each list element represents an individual record. In order to convert this to a [tidy data frame](https://cran.r-project.org/package=tidyr) (in a "tidy" data frame, each variable forms a column, and each record forms a row), the necessary code in base R would be 

```{r}
as.data.frame(t(as.data.frame(z2)))
```  

That is, we first convert it to a data frame, then transpose the result (which converts the data frame to a `"matrix"` object), and then convert it back to a data frame. Using `dmat` is much simpler:

```{r}
dmat(z1, rows = FALSE)  # treat list elements as columns of a data frame
dmat(z2)  # treat list elements as rows of a data frame
```

### Construct block matrices with bmat

Suppose we have three matrices
$$
  A_1 = \begin{pmatrix} 1 & 2 \\ 5 & 6 \end{pmatrix}, \quad
  A_2 = \begin{pmatrix} 3 & 4 \\ 7 & 8 \end{pmatrix}, \quad
  A_3 = \begin{pmatrix} 9 & 10 & 11 & 12 \end{pmatrix}
$$
and want to construct the block matrix defined by
$$
  A = \begin{pmatrix} A_1 & A_2 \\ A_3 \end{pmatrix} = 
    \begin{pmatrix} 1 & 2 & 3 & 4 \\ 
                    5 & 6 & 7 & 8 \\ 
                    9 & 10 & 11 & 12 \end{pmatrix}
$$
In base R, we could accomplish this with the following code:

```{r}
A1 <- matrix(c(1, 2, 5, 6), nrow = 2, byrow = TRUE)
A2 <- matrix(c(3, 4, 7, 8), nrow = 2, byrow = TRUE)
A3 <- matrix(c(9, 10, 11, 12), nrow = 1)
A <- rbind(cbind(A1, A2), A3)
```

This can become rather complicated depending on the structure of the blocks. Specifying the matrices via character strings is more natural and greatly simplifies the task:

```{r, eval=FALSE}
A1 <- mat("1, 2; 5, 6")
A2 <- mat("3, 4; 7, 8")
A3 <- mat("9, 10, 11, 12")
A <- bmat("A1, A2; A3")
```

This function may be familiar to heavy Python+NumPy users since NumPy has a similar function also called `bmat`. See, for example, http://docs.scipy.org/doc/numpy/reference/generated/numpy.bmat.html#numpy.bmat.

## Convenience functions

The main functions in this package are `mat`, `dmat`, and `bmat` and can be viewed as an extension to `matrix`; however, a number of convenience functions are also available, most of which are listed in Table~\ref{tab:functions}. Some of these functions appear in other R packages as well. Of particular note are the [`matlab`](https://cran.r-project.org/package=matlab) and [`pracma`](https://cran.r-project.org/package=pracma) packages.

Function(s)                                | Description
------------------------------------------ | -----------------------------------
`argmax`, `argmin`                         | Find the position of the maximum or minimum in each row or column of a matrix. 
`eye`                                      | Construct an identity matrix. 
`hcat`, `vcat`                             | Concatenate matrices. 
`fill`, `ones`, `zeros`, `trues`, `falses` | Fill a matrix or array with a particular value. 
`flatten`                                  | Flatten or collapse a matrix into one dimension. 
`inv`                                      | Compute the inverse of a square matrix. 
`linspace`, `logspace`                     | Construct a vector of linearly-spaced or logarithmically-spaced elements. 
`meshgrid`                                 | Construct rectangular 2-D grids (useful for plotting). 
`rand`, `randi`, `randn`                   | Construct a matrix or array of pseudorandom numbers. 
`size`, `resize`                           | Extract or change the size and shape of a matrix or array. 
`tr`                                       | Compute the trace of a matrix. 
`tri`, `tril`, `triu`                      | Construct or extract lower and upper triangular matrices.


Although the functionality listed in Table~\ref{tab:functions} can be accomplished in base R (though, not necessarily through a simple function), the `ramify` functions are simple and more familiar to most Julia, MATLAB/Octave, and Python+NumPy users; thus, making the transition to R easier. `linspace` and `logspace` are two such functions. For example, `linspace(1+2i, 10+10i, 8)` creates a vector of complex numbers with 8 evenly spaced points between `1+2i` and `10+10i`. In base R, we would use `seq(1+2i, 10+10i, length = 8)`. There is no base R equivalent to `logspace`. 

Another example is finding the trace of a matrix. To obtain the trace of a matrix in base R the user must manually sum the diagonal entries. Using the `ramify` function `tr` is simpler:

```{r}
m <- rand(10, 10)
sum(diag(m))  # compute the trace in base R
tr(m)  # compute the trace using ramify's tr function
```

The function is called `tr`, rather than `trace`, to avoid conflicts with a base R function called `trace` that is used for interactive tracing and debugging.

Many popular scientific languages also provide convenient ways to generate vectors and matrices of pseudo-random numbers. Commonly found functions are `rand` (for uniform random numbers) and `randn` (for normally distributed random numbers). These and more are available from `ramify`. We saw basic use of `randn` in the section describing the `pprint` function. The following snippet of code compares creating a $100 \times 100 \times 2$ array of $\mathcal{U}\left(0, 1\right)$ random deviates in both base R and `ramify` via the `rand` function.

```{r}
a <- array(runif(20000), c(100, 100, 2))  # base R
a <- rand(100, 100, 2)  # ramify
pprint(a[,,1])  # print the first matrix
```

Of course, the base R approach is more flexible as it allows you to use any one of the built-in distributions (e.g., binomial), however, generating uniform (continuous or discrete) and normal random variates is far more common, hence, the popularity of the `rand`, `randi`, and `randn` functions often seen in other languages.

As seen in the previous example, many of the functions listed in Table~\ref{tab:functions} can be used to construct vectors, matrices, or multi-way arrays just by specifying the extra dimensions. For example,

```{r}
zeros(10)            # 10x1 matrix (i.e., column vector) of zeros
ones(10, 10)         # 10x10 matrix of ones
# fill(pi, 10, 10, 3)  # 10x10x3 array filled with the constant pi
```

Note that these functions, by default, always return a `"matrix"` or `"array"` object meaning there is always a `"dim"` attribute. In contrast, in base R, vectors do not have a `"dim"` attribute. So, when creating a vector using `zeros(10)`, for example, the result will be a matrix of zeros with ten rows and one column. To bypass this behavior, you can use the `atleast_2d` option:

```{r}
zeros(10, atleast_2d = FALSE)  # has class "integer", rather than "matrix"
```

This behavior can also be changed globally using `options(atleast_2d = FALSE)`.

As a last example, we shal demonstrate the `meshgrid` function. A `meshgrid` function is available in MATLAB/Octave and Python+NumPy and is most frequently used to produce input for a 2-D or 3-D function that will be plotted. It should be noted, however, that the R version of `meshgrid` provided by `ramify` returns a list of matrices. The following code, for example, plots contours of the function
$$
  y = \cos{\left(x_1^2 + x_2^2\right)} \times \exp\left(-\frac{\sqrt{x_1^2 + x_2^2}}{6}\right)
$$
over the Cartesian grid $\left[-4\pi, 4\pi\right] \times \left[-4\pi, 4\pi\right]$. The resulting plot is created as follows.

```{r, fig.cap='Color image and contour lines for $y$.', fig.width=7, fig.height=5}
x <- meshgrid(linspace(-4*pi, 4*pi, 27))  # list of input matrices
y <- cos(x[[1]]^2 + x[[2]]^2) * exp(-sqrt(x[[1]]^2 + x[[2]]^2)/6)
par(mar = c(0, 0, 0, 0))  # remove margins
image(y, axes = FALSE)  # color image
contour(y, add = TRUE, drawlabels = FALSE)  # add contour lines
```

## Summary

Matrices are a fundamental feature of any scientific language. This vignette introduced the simple R package `ramify`, which provides additional matrix functionality in R. Using `ramify`, I showed how to craft matrices using: (i) character strings and (ii) lists. A similar construction of block matrices and data frames was also briefly discussed. I also provided a quick summary of a number of convenience functions for easing the transition to R from other popular scientific languages such as Julia, MATLAB/Octave, and Python+NumPy.
